Lessons learned: creating tutorials to teach agent-based modelling to archaeologists

### 2025-06-20 paper.Rmd
update paper

Agent-based modelling (ABM) is one of the most widely applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists. However, the access to vocational training for ABM is limited. In this paper we present the Open Educational Resources (OER) that we produced in an international cooperation. For this purpose we created online and open tutorials for ABM for archaeologists. In addition, how-to guides and support material were created, both for teachers and students. The materials adhere to international (European) standards in relation to education and the Digital Skills Passport for archaeology. The teaching materials have been well tested using a diverse group of archaeologists during various international conferences and meetings. It was shown that the knowledge and skills of the participants in relation to ABM improved. In addition, the demand for the teaching materials and the workshops was high, ensuring that the tutorials on ABM will be used in the future.

@@

The ABMA project sought to create Open (Open Access and Open Source) Educational Resources (OERs) for training in Agent-Based Modelling for Archaeologists. These OERs were to be HTML- and JavaScript-based so they could be used on any device with a browser, even if it is not connected to the internet. The full programme consists of five hands-on practical tutorials, taking learners through the process of creating the digital models. The modular architecture ensures that they can be independent or incorporated into other teaching tools like learning management systems. They are also deliberately 'method agnostic' so they can be incorporated into any tools from self-teaching to MOOCs, and are supported by a set of materials to encourage use of the OERs.

@@

1.  a surge in professionals who can build, apply and critique archaeological simulation, thus opening new research avenues for the discipline as a whole;

2.  increased level of digital competence, specifically in advanced computer skills that are highly transferable to other industries. This includes technical skills such as programming but also "soft" skills related to computational thinking;

@@

Over the course of the project three groups of students worked on the project during the Smart Solutions Semester at Saxion University of Applied Sciences. This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (<https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions>). Within the semester students have to come up with new ideas or solutions to help a client with a wicked, or at least challenging, problem. The backgrounds of the various students we worked with were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media & Game Technologies, and ICT. The aim of involving interdisciplinary groups in the project was to involve students in current research projects, to enable them to learn new things and share their innovative ideas and solutions with us. The outcomes of the students groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project. Their help during workshops was useful, but they also were engaged in developing educational material [@aalpoel2024], testing the tutorials, and developing a style for the website [@jutte2024] and other materials.

@@

Each tutorials consists of a number of individual lessons, which guide the learner in a self-paced manner through an increasingly challenging set of exercises. Collectively, they build the skills and knowledge necessary to become a confident agent-based modeller.

In the first tutorial the users learn what simulations and agent-based models are and how they can help in archaeological research (Figure \@ref(fig:screenshot-lesson-3)). This tutorial consists of four lessons and makes use of the Artificial Anasazi model [@axtell2002]. The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific. Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants take their first steps in NetLogo syntax. The third lesson explains how ABMs are used in archaeological research. The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode. This tutorial builds a foundation for working in NetLogo and with ABM. All activity stays within the entry level of difficulty, although some higher level concepts are lightly touched upon.

@@

In the fifth and final tutorial the learner learns more about how to incorporate agent-based modelling in archaeological research. This tutorial focuses less on programming in NetLogo and more on the principle and standards of computer-based modelling. The tutorial consists of five lessons. It summarizes the material delivered in the previous lessons and relates it to the phases of the model development process: the conceptual phase, the technical phase and the dissemination phase. This tutorial approaches more theoretical aspects in a practical environment. For the conceptual phase, particular focus is directed at a discussing what kind of research questions are suitable for modelling and how to pick the right modelling technique. The importance of properly conceptualizing a model before starting the technical part of model development is highlighted as it is a common issue for less experienced modellers. The experiment design phase, including parametrisation, validation and the analysis and interpretation of the models is explained at length since it often proves challenging for students. On a practical side, the BehaviorSpace - NetLogo's experiment environment - is explained to enable the learners to batch run their models. Finally the dissemination phase of the model development is also included to ensure that the standards in model publication are widely known and the learners explore the fundamental importance of replication for advancing scientific knowledge.

@@

A large proportion of the participants of the workshops gave us information using the survey before the workshops (172 of 368 participants). The respondents came from at least 40 countries that represented almost all continents (see Figure \@ref(fig:nationality)). There was a clear skew towards participants from Europe. The number of female respondents slightly outnumbered the male ones and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)). The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years of age, indicating that the audience was predominantly composed of early career archaeologists. It seems that the different events also had a slightly different distribution of both gender and age. For example, more older people attended the workshop at the CAA conference in April 2023 and more men were present during the workshop at the Reuvensdagen in November 2023. This might be due to the differences of audiences at the conferences. For the Reuvensdagen the number of participants of the workshop was relatively low and this variance might be due to chance.

@@

During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials (Figure \@ref(fig:workshop-impression)). Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research or they discussed possibilities for the application of ABM in general. Individual progress varied depending on skills and preferences of the participants. While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.

@@

A similar proportion of the participants of the workshops gave us feedback after the workshops (171 of 368 participants).

@@

Most respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good. The teachers were rated even better than the teaching materials, with most respondents (165) giving them the *excellent* score (see Figure \@ref(fig:rating-teaching)).

@@

For *Competence area 1. Information and data literacy*, the tutorials mainly address *1.2 Evaluating Data, Information And Digital Content* and *1.3 Managing Data, Information And Digital Content*. In the ABMA educational materials, users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM. For tutorial 2 the user is expected to be able to find information using the NetLogo(Web) dictionary (<http://ccl.northwestern.edu/netlogo/bind/> or <http://ccl.northwestern.edu/netlogo/docs/dictionary.html>).

@@

*Competence Area 3. Digital content creation* is one of the main foci of the ABM teaching material. In the process of learning ABM the users are constantly *3.1 Developing Digital Content* and *3.2 Integrating And Re-Elaborating Digital Content*. The material does not touch upon *3.3 Copyright And Licences*, as they fall out of the scope of this project. A very important aspect of learning ABM is *3.4 Programming*. The users are guided through the process of developing the necessary programming skills to build a simulation. While the programming language used - NetLogo - is specific to ABM, its syntax follows general programming patterns and the learning is thus transferable to other programming languages. The users are expected to learn both syntax and more general computational concepts such as modular code development, testing and documenting code.

*Competence Area 5. Problem solving* is very important and is closely tied to *3.4 Programming*, since programming involves a lot of problem solving but also the general principles of model development that have a prominent place in the educational materials. In addition, the creation of a highly complex ABM consists of problem solving all the time, including development of research questions, conceptualisation of theories, experiment design and validating against data - all of which require creative abilities and a high level of understanding the scientific process. All competencies are addressed: *5.1 Solving Technical Problems*, is common when writing code, learning about coding, error handling and going from pseudocode to real code. The competence *5.2 Identifying Needs And Technological Responses*, is a central issue of the course, while *5.3 Creatively Using Digital Technology* is an important aspect of so-called model thinking necessary to develop robust and insightful models. In tutorial 2 various aspects of *5.4 Identifying Digital Competence Gaps*, are relevant, since this tutorial forces the user to assess their knowledge and skills.

@@

A digital skills passport is the same concept as a physical skills passport but it is digitised to overcome some of the drawbacks of a physical skills passport:

@@

The chosen workflow for developing the tutorials, namely testing the tutorials ourselves and then testing them during workshops and conferences proved very valuable in many ways. First of all, it had a very practical use; new users found bugs that we as a team missed or did not encounter. Participants used different operating systems, different web browsers and different settings on their computers, leading to other uses and other bugs. In addition, different people interact differently with software on their computers. These behavioural differences between users helped us to understand how people would use our tutorials and enabled improvements. Thirdly, our presence at conferences and the organization of workshops improved dissemination of the tutorials. A fourth aspect is related, since the high number of people interested in these workshops has shown the need and demand for educational materials for ABM.

@@

Based on the high demand for our online and offline workshops, it is possible to conclude that "resistance to digital tools" [@morgan2022, 216] might be dwindling. The demand for our tutorials and workshops, shows that there is an increasing group of archaeologists willing to learn a new tool. At the same time the tutorials help to move ABM away from the fringes of archaeology by offering an easy access to ABM and programming. However, we also noted that some participants found the workshops interesting, but would not pursue further learning about ABM. Although, these participants will probably never apply ABM in their research, they are now aware of the possibilities of the method, at least acknowledging the possibilities of ABM and digital archaeology in the broader sense.

Teaching digital archaeology is often contingent on the skills and interest of the teaching staff of a gavine higher education institute, although some institutes are incorporating digital aspects of archaeology in their regular curriculum [@visser2016; @lambers2024]. The number of specialists in ABM is limited world wide and the OER's created in this project enable archaeologists to provide an easy introduction into ABM, modelling and computational thinking. However, we encourage people to discuss their ideas with other people in the field, for example during CAA conferences (<https://caa-international.org/>). We have seen during the workshops that discussion is probably the most important aspect of model building, which is impossible to incorporate in self-paced learning.

@@

Future prospects of these open learning materials will be the further application and integration in the educational programmes in our universities and hopefully in others too. We have made the material openly available and we hope that the ABM-community will stay involved and that future updates will be community driven.

>>>>

Agent-based modelling (ABM) is one of the most widely applied computational modelling research techniques in archaeology, because it is well suited to the questions asked by archaeologists. However, the access to vocational training for ABM is limited. In this paper we present the Open Educational Resources (OER) that we produced in an international cooperation. For this purpose, we created online and open tutorials for ABM for archaeologists. In addition, how-to guides and support material were created, both for teachers and students. The materials adhere to international (European) standards in relation to education and the Digital Skills Passport for archaeology. The teaching materials have been well tested using a diverse group of archaeologists during various international conferences and meetings. It was shown that the knowledge and skills of the participants in relation to ABM improved. In addition, the demand for the teaching materials and the workshops was high, ensuring that the tutorials on ABM will be used in the future.

@@

The ABMA project sought to create Open (Open Access and Open Source) Educational Resources (OERs) for training in Agent-Based Modelling for Archaeologists. These OERs were to be HTML- and JavaScript-based so they could be used on any device with a browser, even if it is not connected to the internet. The full programme consists of five hands-on practical tutorials, taking learners through the process of creating digital models. The modular architecture ensures that they can be independent or incorporated into other teaching tools like learning management systems. They are also deliberately 'method agnostic' so they can be incorporated into any tools from self-teaching to MOOCs, and are supported by a set of materials to encourage use of the OERs.

@@

1.  a surge in professionals who can build, apply, and critique archaeological simulation, thus opening new research avenues for the discipline as a whole;

2.  increased level of digital competence, specifically in advanced computer skills that are highly transferable to other industries. This includes technical skills such as programming, but also "soft" skills related to computational thinking;

@@

Over the course of the project three groups of students worked on the project during the Smart Solutions Semester at Saxion University of Applied Sciences. This is an interdisciplinary semester in which students of at least three different study-programmes or disciplines work together on a complex problem/project (<https://www.saxion.edu/business-and-research/collaborate-with-saxion/smart-solutions>). Within the semester students have to come up with new ideas or solutions to help a client with a wicked, or at least challenging, problem. The backgrounds of the various students we worked with were diverse: Applied Computer Science, Archaeology, Business Management Studies, Creative Business, Creative Media & Game Technologies, and ICT. The aim of involving interdisciplinary groups in the project was to involve students in current research projects, to enable them to learn new things and share their innovative ideas and solutions with us. The outcomes of the student groups were supplementary to the deliverables of the project, but their new ideas were beneficial to the project. Their help during workshops was useful, but they also were engaged in developing educational material [@aalpoel2024], testing the tutorials, and developing a style for the website [@jutte2024] and other materials.

@@

Each tutorial consists of a number of individual lessons, which guide the learner in a self-paced manner through an increasingly challenging set of exercises. Collectively, they build the skills and knowledge necessary to become a confident agent-based modeller.

In the first tutorial the users learn what simulations and agent-based models are and how they can help in archaeological research (Figure \@ref(fig:screenshot-lesson-3)). This tutorial consists of four lessons and makes use of the Artificial Anasazi model [@axtell2002]. The first two lessons introduce the learner to simulation in general and Agent-Based Modelling in specific. Various concepts related to ABM, such as the definitions of model and simulation, are introduced and the participants take their first steps in NetLogo syntax. The third lesson explains how ABMs are used in archaeological research. The final lesson introduces the learner to the NetLogo Interface and the difference between the Interactive and Authoring mode. This tutorial builds a foundation for working in NetLogo and with ABM. All activity stays within the entry level of difficulty, although some higher-level concepts are lightly touched upon.

@@

In the fifth and final tutorial the learner learns more about how to incorporate agent-based modelling in archaeological research. This tutorial focuses less on programming in NetLogo and more on the principle and standards of computer-based modelling. The tutorial consists of five lessons. It summarizes the material delivered in the previous lessons and relates it to the phases of the model development process: the conceptual phase, the technical phase and the dissemination phase. This tutorial approaches more theoretical aspects in a practical environment. For the conceptual phase, particular focus is directed at discussing what kind of research questions are suitable for modelling and how to pick the right modelling technique. The importance of properly conceptualizing a model before starting the technical part of model development is highlighted as it is a common issue for less experienced modellers. The experiment design phase, including parametrisation, validation and the analysis and interpretation of the models is explained at length since it often proves challenging for students. On a practical side, the BehaviorSpace - NetLogo's experiment environment - is explained to enable the learners to batch run their models. Finally, the dissemination phase of model development is also included to ensure that the standards in model publication are widely known, and the learners explore the fundamental importance of replication for advancing scientific knowledge.

@@

A large proportion of the participants in the workshops gave us information using the survey before the workshops (172 of 368 participants). The respondents came from at least 40 countries that represented almost all continents (see Figure \@ref(fig:nationality)). There was a clear skew towards participants from Europe. The number of female respondents slightly outnumbered the male ones, and a small group did not share their gender, while two identified as non-binary (see Figure \@ref(fig:gender-age)). The ages of the respondents ranged from below 20 to over 70, although the majority fell between 20 and 40 years of age, indicating that the audience was predominantly composed of early career archaeologists. It seems that the different events also had a slightly different distribution of both gender and age. For example, more older people attended the workshop at the CAA conference in April 2023, and more men were present during the workshop at the Reuvensdagen in November 2023. This might be due to the differences of audiences at the conferences. For the Reuvensdagen the number of participants of the workshop was relatively low and this variance might be due to chance.

@@

During the workshops the participants worked in a self-paced manner, often on their own, but sometimes working together and discussing the tutorials (Figure \@ref(fig:workshop-impression)). Some participants engaged in discussion with the teachers to learn how they could apply ABM in their research, or they discussed possibilities for the application of ABM in general. Individual progress varied depending on the skills and preferences of the participants. While bugs were still a problem in earlier workshops, participant-teacher interaction was more focused on content-related learning problems in later installments.

@@

A similar proportion of the participants in the workshops gave us feedback after the workshops (171 of 368 participants).

@@

Most respondents (153) were enthusiastic about the teaching materials, with the majority rating them as excellent or good. The teachers were rated even better than the teaching materials, with most respondents (165) giving them an *excellent* score (see Figure \@ref(fig:rating-teaching)).

@@

For *Competence area 1. Information and data literacy*, the tutorials mainly address *1.2 Evaluating Data, Information and Digital Content* and *1.3 Managing Data, Information And Digital Content*. In the ABMA educational materials, users learn about working with data, being critical of their digital content and how to model and manage the data in the context of an ABM. For tutorial 2 the user is expected to be able to find information using the NetLogo(Web) dictionary (<http://ccl.northwestern.edu/netlogo/bind/> or <http://ccl.northwestern.edu/netlogo/docs/dictionary.html>).

@@

*Competence Area 3. Digital content creation* is one of the main foci of the ABM teaching material. In the process of learning ABM the users are constantly *3.1 Developing Digital Content* and *3.2 Integrating and Re-Elaborating Digital Content*. The material does not touch upon *3.3 Copyright and Licences*, as they fall out of the scope of this project. A very important aspect of learning ABM is *3.4 Programming*. The users are guided through the process of developing the necessary programming skills to build a simulation. While the programming language used - NetLogo - is specific to ABM, its syntax follows general programming patterns and the learning is thus transferable to other programming languages. The users are expected to learn both syntax and more general computational concepts such as modular code development, testing and documenting code.

*Competence Area 5. Problem solving* is very important and is closely tied to *3.4 Programming*, since programming involves a lot of problem solving but also the general principles of model development that have a prominent place in the educational materials. In addition, the creation of a highly complex ABM consists of problem solving all the time, including development of research questions, conceptualisation of theories, experiment design and validating against data - all of which require creative abilities and a high level of understanding the scientific process. All competencies are addressed: *5.1 Solving Technical Problems*, is common when writing code, learning about coding, error handling and going from pseudocode to real code. The competence *5.2 Identifying Needs and Technological Responses*, is a central issue of the course, while *5.3 Creatively Using Digital Technology* is an important aspect of so-called model thinking necessary to develop robust and insightful models. In tutorial 2 various aspects of *5.4 Identifying Digital Competence Gaps*, are relevant, since this tutorial forces the user to assess their knowledge and skills.

@@

A digital skills passport is the same concept as a physical skills passport, but it is digitised to overcome some of the drawbacks of a physical skills passport:

@@

The chosen workflow for developing the tutorials, namely testing the tutorials ourselves and then testing them during workshops and conferences proved very valuable in many ways. First of all, it had a very practical use; new users found bugs that we as a team missed or did not encounter. Participants used different operating systems, different web browsers and different settings on their computers, leading to other uses and other bugs. In addition, different people interact differently with software on their computers. These behavioural differences between users helped us to understand how people would use our tutorials and enabled improvements. Thirdly, our presence at conferences and the organization of workshops improved dissemination of the tutorials. A fourth aspect is related, since the high number of people interested in these workshops has shown the need and demand for educational materials for ABM and to improve computational skills.

@@

Based on the high demand for our online and offline workshops, it is possible to conclude that "resistance to digital tools" [@morgan2022, 216] might be dwindling. The high demand was striking in Amsterdam during CAA 2023. The large number of registered participants forced us to expand the workshop and to split the group in two different rooms to account for all people. In addition, the number of registrations for the online workshop was far exceeding our expectations. While not everyone showed up, this does show that there is a high interest and demand. During the various workshops we encountered many people wanting to apply ABM to their own research, but lacking the skills or tools. The high demand for our tutorials and workshops shows that there is an increasing group of archaeologists willing to learn about new tools. At the same time, the tutorials help to move ABM away from the fringes of archaeology by offering an easy access to ABM and programming. Of course, some participants who found the workshops interesting may still not pursue ABM further. However, while they may not end up applying ABM in their own research, they are now more aware of the possibilities of ABM and digital archaeology more generally.

Teaching digital archaeology is often contingent on the skills and interest of the teaching staff of a given higher education institute, although some institutes are incorporating digital aspects of archaeology in their regular curriculum [@visser2016; @lambers2024]. The number of specialists in ABM is limited worldwide and the OER's created in this project offers archaeologists an easy introduction into ABM, modelling and computational thinking. However, we encourage people to discuss their ideas with other people in the field, for example during CAA conferences ([https://caa-international.org/](#0){.uri}). We have seen during the workshops that discussion can ben an important aspect of model building, one which is challenging to incorporate in self-paced learning.

@@

Future prospects of these open learning materials will be the further application and integration in the educational programmes in our universities and hopefully in others too. It would be interesting to test the effectiveness of the OERs in the future. We have made the material openly available. We hope that the ABM-community will stay involved and that future updates will be community driven.
